\chapter{Where do we go from here?} \label{final conclusion}

The present book set out to investigate the interface of morphology, \isi{phonology and phonetics} by investigating \isi{gemination} in English \isi{affixation}. 
One main aim of this book was to clarify the role of \isi{boundary strength} on the \isi{acoustic realization} of complex words. 
While previous studies have found evidence for effects of \isi{boundary strength} on the phonetics of words, the specific nature of these effects is yet unclear (cf. chapter \ref{Gemination}). 
 Clarifying the nature of these effects is, however, quite important, as a clarification is necessary to accurately model the morpho-phonological and the \isi{morpho-phonetic interface}. 

As shown in chapter \ref{Theory}, different formal linguistic and psycholinguistic theories deviate in their conceptualization of  \isi{boundary strength}. While some theories assume a categorical difference between sets of affixes (e.g. \isi{Lexical Phonology}, Stratal OT), others assume \isi{boundary strength} to be a gradient, probabilistic word-specific concept (e.g. the Decomposability Approach, the Morphological Informativeness Approach). 
While some approaches define \isi{boundary strength} by means of mainly lexical factors (e.g. \isi{Lexical Phonology}, Stratal OT), others mainly focus on prosodic aspects (e.g. Prosodic Phonology), and others concentrate on semantics (e.g. Morphological Informativeness).  The differences in the conceptualization of \isi{boundary strength} mirror general differences in theoretical assumptions about the \isi{morpho-phonological interface}. % This mirrors general differences in theoretical assumptions about the morpho-phonological and the \isi{morpho-phonetic interface}. 
 The different conceptualizations of \isi{boundary strength} also lead to different predictions for the \isi{phonetic realization} of complex words.
 Studies which test these predictions will, in turn, have important implications about the nature of the \isi{interface between morphology}, \isi{phonology and phonetics}.
 
To further investigate possible effects of \isi{boundary strength}, and to investigate the nature of the morpho-phonological and the \isi{morpho-phonetic interface}, I conducted a corpus study and an experimental study on \isi{morphological gemination} in English.  As morphological \isi{geminates} always occur across morphological boundaries, they provide the perfect test case for investigating possible effects of \isi{boundary strength}. 
 In the studies, I investigated the five English affixes \prefix{un}, negative \prefix{in}, locative \prefix{in}, \prefix{dis} and \suffix{ly}. I tested the predictions of various morpho-phonological and morpho-phonetic approaches. By finding out which approach can account best for the variation in \isi{gemination} with English affixes, important implications about the \isi{interface between morphology}, \isi{phonology and phonetics} can be drawn.



 
The studies revealed that while the prefix \prefix{un} \isi{geminates} in corpus and \isi{experimental speech}, the prefixes locative \prefix{in} and negative \prefix{in} show differences in their \isi{gemination} pattern depending on \isi{speech mode}. In the corpus data, in which one can assume a deeper \isi{semantic processing} than in the experimental data, the prefixes \isi{geminate} to a similar degree as \prefix{un}. In the experimental data, they show smaller durational differences between doubles and singletons than \prefix{un}, and \isi{gemination} depends on the prosody of the derivatives. The prefix \prefix{dis} \isi{geminates} to a weaker degree than \prefix{un} in both studies, and the suffix  \suffix{ly} never \isi{geminates}.

These results falsify common assumptions about \isi{gemination} in English (cf. \sectref{Gemination in English}). They also falsify \isi{stratal approaches} of the \isi{morpho-phonological interface} (e.g. \citealt{Kiparsky.1982,Kiparsky.1985,Mohanan.1986,BermudezOtero.2012,Kiparsky.2015,BermudezOtero.2017}). 
The variation in \isi{gemination} can best be accounted for by the \isi{morphological informativeness} and semantic \isi{segmentability} of the affixes. The more meaning an affix carries and the more informative it is, the stronger it \isi{geminates}. 

The results, furthermore, indicate that, at least in some cases, prosodic structure affects \isi{gemination}. It is, however, yet to be investigated how the \isi{informativeness} and \isi{segmentability} of an affix relate to prosodic structure, and how this relation affects the \isi{acoustic realization} of complex words. 
While there are already approaches which assume a close relation between the prosodic structure of a word and its \isi{segmentability}, such as the Prosodic Word Approach proposed by \cite{Raffelsiefen.1999}, these approaches are currently not specified enough to account for the findings of this book.  More research on the interaction between the prosodic structure of a complex word, its \isi{segmentability} and its  \isi{informativeness} is needed to devise a more adequate model of the \isi{morpho-phonological interface}.

Apart from \isi{gemination}, the studies also revealed general effects of \isi{decomposability} on the \isi{acoustic realization} of complex words. These effects have important implications for models of \isi{morphological processing} and \isi{models of speech production}. It was shown that both categorical affix-specific \isi{segmentability} and gradient \isi{word-specific decomposability} affect the \isi{acoustic realization} of complex words. Theories of morpho-phonological processing must thus be devised to account for categorical and gradient effects of morphological structure on speech production.
Furthermore, the corpus study revealed that the two \isi{homophonous} prefixes negative \prefix{in} and locative \prefix{in} were realized with different nasal durations. This result calls for \isi{speech production models} which allow morphological structure, or some correlate thereof, to affect \isi{phonetic detail}. 

To conclude, the present book shows that current approaches of the morpho-phonological and \isi{morpho-phonetic interface} are not able to account satisfactorily for the variation found in the \isi{phonetic realization} of complex words. 
 The studies conducted provide empirical evidence that there is a close connection between between \isi{morphological informativeness}, \isi{segmentability} and prosody, and that these three factors are crucial in the production of complex words. 
Further research is needed to accurately model their effects on the \isi{phonetic realization} of complex words.
Furthermore, the results of this book call for a revision of \isi{speech production models}. These models must specify the \isi{morpho-phonological-phonetic interface} in such a way that it is able to account for categorical and gradient effects of morphological structure on the \isi{acoustic realization} of complex words.
